---
title: "examine_contributing_species_PCoA"
author: "Ryan Rebernick"
date: "6/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description:
This looks at the output of the corr.axes function of mothur (https://mothur.org/wiki/corr.axes/#:~:text=Default%20settings-,The%20corr.,temperature%2C%20weight%2C%20etc.)) in order to determine which OTUs are responsible for shifting the samples along the 2 axes. We determine this by measuring the correlation of the relative abundance of each OTU with the two axes using the corr.axes function. Used on PCoA plots from mothur. 
This script merges the output of corr.axes with the OTU taxonomy infomration from mothur so we know the identities of the OTUs impacting the axes most. 

# Load packages
```{r, warning=F}
library(dplyr)
library(magrittr)
library(data.table)
library(ggfortify)
library(cowplot)


```


# Examine differences for PCoA1/PCoA2

## View PCoA plot

### Late time point only
```{r}

axes <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_otu.thetayc.0.03.lt.ave.pcoa.axes')
loadings <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_otu.thetayc.0.03.lt.ave.pcoa.loadings')
geno <- fread('~/Desktop/projects/reuterin/data/mothur/SAMPLE-GENOTYPE.txt')

# combine
pcoa <- merge(axes, geno, by.x = 'group', by.y='V1') %>%
  dplyr::mutate(time = ifelse(stringr::str_detect(group, 'S'), 'Late', 'Early')) %>%
  dplyr::mutate(cage = gsub('\\..*', '', group)) %>%
  dplyr::select(group, V2, time, cage, axis1:axis10) %>%
  set_colnames(c('sample', 'Genotype', 'Time', 'Cage', paste0('PCoA', 1:10))) %>%
  dplyr::filter(Time == 'Late')

# remove sample 516.1S
pcoa %>%
  dplyr::filter(sample != '516.1S') %>%
  ggplot(., aes(x=PCoA1, y=PCoA2, color=Genotype)) +
  geom_point(size=6) +
  xlab(paste0('PCoA1 (', round(loadings$loading[1], 1), '%)')) +
  ylab(paste0('PCoA2 (', round(loadings$loading[2], 1), '%)')) +
  ggtitle(label = 'Late timepoints') +
  theme_cowplot(12) +
  theme(title =  element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20)) +
  scale_color_viridis(discrete = T, end = 0.92)


```



# Combine files showing what OTUs are and which OTUs contribute to PCoA axes for late stage samples

## Combine files
```{r}

otu <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_A.pick.opti_mcc.0.03.cons.taxonomy', data.table = F)
corr.axes <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_late_outRem.spearman.corr.axes', data.table = F) %>%
  merge(., otu, by = 'OTU') %>%
  set_colnames(c('OTU', 'pcoa1', 'pcoa1_p', 'pcoa2', 'pcoa2_p', 'length', 'size', 'taxonomy'))
corr.axes %<>% arrange(-length)

write.table(corr.axes, '~/Desktop/projects/reuterin/int/pcoa_late_contributing_otus.txt', row.names = F, sep = '\t', quote = F)


```



# Map length vectors onto PCoA (late time point, outliers removed)

## Plot
```{r}

axes <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_otu.thetayc.0.03.lt.ave.pcoa.axes')
loadings <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_otu.thetayc.0.03.lt.ave.pcoa.loadings')
geno <- fread('~/Desktop/projects/reuterin/data/mothur/SAMPLE-GENOTYPE.txt')

# combine
pcoa <- merge(axes, geno, by.x = 'group', by.y='V1') %>%
  dplyr::mutate(time = ifelse(stringr::str_detect(group, 'S'), 'Late', 'Early')) %>%
  dplyr::mutate(cage = gsub('\\..*', '', group)) %>%
  dplyr::select(group, V2, time, cage, axis1:axis10) %>%
  set_colnames(c('sample', 'Genotype', 'Time', 'Cage', paste0('PCoA', 1:10))) %>%
  dplyr::filter(Time == 'Late')

# remove sample 516.1S
pcoa %>%
  dplyr::filter(sample != '516.1S') %>%
  ggplot(., aes(x=PCoA1, y=PCoA2, color=Genotype)) +
  geom_point(size=6) +
  xlab(paste0('PCoA1 (', round(loadings$loading[1], 1), '%)')) +
  ylab(paste0('PCoA2 (', round(loadings$loading[2], 1), '%)')) +
  ggtitle(label = 'Late timepoints') +
  theme_cowplot(12) +
  theme(title =  element_text(size=20),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20)) +
  scale_color_viridis(discrete = T, end = 0.92)+ 
  geom_segment(aes(x = 0, y = 0, xend = corr.axes$pcoa1[1], yend = corr.axes$pcoa2[1]), arrow = arrow(length = unit(5, 'points')), color='black')




```





# Combine metastats output and export

## Load metastats data for late time point, outlier removed
non-parametric T-tetst that determines whether there are any OTUs that are differentially represented between the samples 
Need to correct p values for multiple hypothesis testing per (https://forum.mothur.org/t/multiple-corrections-and-unique-sequences/20086)
Group 1 = First name in file name
Group 2 = 2nd name in file name
```{r}

# otu information
otu <- fread('~/Desktop/projects/reuterin/data/mothur/reuterin_A.pick.opti_mcc.0.03.cons.taxonomy', data.table = F)
# metastats output
## comparison 1
metastats <- fread('~/Desktop/projects/reuterin/data/mothur/metastats/reuterin_late_outRem.0.03.TripleMut_CRE_NEG.metastats', data.table = F) %>%
  merge(., otu, by= 'OTU') %>%
  set_colnames(c('OTU', 'tripleko_mean', 'tripleko_var', 'tripleko_SE', 'cre_mean', 'cre_var', 'cre_SE', 'pval_3ko.cre', 'size', 'taxonomy'))
## comparison 2
metastats <- fread('~/Desktop/projects/reuterin/data/mothur/metastats/reuterin_late_outRem.0.03.DoubleMut_CRE_NEG.metastats', data.table = F) %>%
  set_colnames(c('OTU', 'doubleko_mean', 'doubleko_var', 'doubleko_SE', 'cre_mean', 'cre_var', 'cre_SE', 'pval_2ko.cre')) %>%
  merge(metastats, ., by = 'OTU')
## comparison 3
metastats <- fread('~/Desktop/projects/reuterin/data/mothur/metastats/reuterin_late_outRem.0.03.TripleMut_DoubleMut.metastats', data.table = F) %>%
  set_colnames(c('OTU', 'tripleko_mean', 'tripleko_var', 'tripleko_SE', 'doubleko_mean', 'doubleko_var', 'doubleko_SE', 'pval_3ko.2ko')) %>%
  merge(metastats, ., by = 'OTU')

# clean 
metastats %<>%
  dplyr::select(OTU, taxonomy, size, cre_mean.x, cre_SE.x, doubleko_mean.x, doubleko_SE.x, tripleko_mean.x, tripleko_SE.x, pval_2ko.cre, pval_3ko.cre, pval_3ko.2ko) %>%
  set_colnames(gsub('\\.[x|y]', '', colnames(.)))

# calculate adjusted pvalues
metastats %<>% 
  dplyr::mutate(adj.pval_2ko.cre = p.adjust(pval_2ko.cre, method = 'BH')) %>%
  dplyr::mutate(adj.pval_3ko.cre = p.adjust(pval_3ko.cre, method = 'BH')) %>%
  dplyr::mutate(adj.pval_3ko.2ko = p.adjust(pval_3ko.2ko, method = 'BH')) %>%
  arrange(pval_3ko.cre,pval_2ko.cre,pval_3ko.2ko)

# save
write.table(metastats, '~/Desktop/projects/reuterin/out/tables/metastats_late_outRem_combined.txt', quote = F, sep = '\t', row.names = F, col.names = T)

```



