---
title: "1_prepareForCytoscape"
author: "Ryan Rebernick"
date: "2/26/2021"
output: html_document
---

# Load packages
```{r, warning=F, message=F}

library('plyr')
library('dplyr')
library('magrittr')
library('data.table')

#BiocManager::install("biomaRt")
library('biomaRt')
library('ggplot2')
library('scales')
library(cowplot)

```



# Prepare Metabolite data
Input data was obtained through the following process:
- Took list of metabolites shown to be changed following treatment with lactobacillus reuteri in SW480 and HCT cells
- Converted the common names of these metabolites to HMDB files using the GUI version of metaboanalyst v5.0 (https://www.metaboanalyst.ca/faces/upload/ConvertView.xhtml). This was done via GUI because a fair number of the compounds were not directly identified and had to be selected from a list of suspected matches.
- Once these HMDB ids were obtained, MetaBridge v1.2 (https://www.metabridge.org/) was used to get lists of genes associated with each compound - these were mapped using KEGG and MetaCyc. Both values were taken (see below)

The below code combines the kegg and metacyc hits for the HCT and SW cell lines, selecting for variables needed.
```{r}

#  Load metabridge data
kegg <- fread('~/Desktop/projects/reuterin/int/metabolites_SwHctSig_HMDB_metaboAnalyst_5.0_mapped_KEGG.tsv', data.table = F)
met <- fread('~/Desktop/projects/reuterin/int/metabolites_SwHctSig_HMDB_metaboAnalyst_5.0_mapped_MetaCyc.tsv', data.table = F)

# Prepare for merging
kegg$cg <- paste0(kegg$HMDB, '-', kegg$`Gene Name`)
met$cg <- paste0(met$HMDB, '-', met$`Gene Name`)

# Merge data frames
metabridge <- merge(kegg, met, by= 'cg')

# select values needed
metabridge$interaction <- 'mg'
metabridge %<>% dplyr::select(Compound.x, interaction, `Gene Name.y`, KEGG.x:`Enzyme Name`, Reaction:`MetaCyc Gene`, Ensembl)
colnames(metabridge) <- c('compound', 'interaction', 'gene', 'KEGG', 'HMDB', 'Enzyme', 'enzyme_name', 'rxn', 'rxn_name', 'metaCyc_gene', 'ensembl')

# shorten for export to cytoscape
export <- metabridge %>% dplyr::select(compound:gene) %>% unique()

```



# Add Proteomics data
Cysteine proteomics data was done at the Broad to look at cysteines preferentially bound by reuterin. The lab recommended a cutoff of 2, but as this yielded only 3 compounds instead a value of >Mean + 3SD in any of the cr values. My understanding is the CR values are basically a ratio comparing reuterin and a random compound. The higher the ratio the more reuterin binds that cysteine site.  

### Initial Cysteine proteomics analysis 
Filter for CR ratios in reuterin > Mean+3SD. From there if multiple cystieines are significant for each gene, take the highest cpmax value for each protein. This will be combined with the subsequent cysteine proteomics analysis d/t the high variability in protein capture between experiments. 
```{r}

# load and rename cols
prot <- fread('~/Desktop/projects/reuterin/data/UMichigan_Hannah_HCT_Cys_SLC-ABPP.txt', data.table = F)
colnames(prot)[28:32] <- c('cr_acro_5', 'cr_acro_25', 'cr_reuterin_50', 'cr_reuterin_100', 'gene-site')

# Correct excel's goofy date switches
prot$gene_symbol <- ifelse(prot$gene_symbol == '10-Sep', 'SEP10', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '9-Sep', 'SEPT9', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '11-Sep', 'SEP11', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '2-Sep', 'SEPT2', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '5-Sep', 'SEPT5', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '7-Sep', 'SEPT7', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '8-Sep', 'SEPT8', prot$gene_symbol)

prot$gene_symbol <- ifelse(prot$gene_symbol == '2-Mar', 'MOSC2', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '5-Mar', 'MARH5', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '7-Mar', 'MARH7', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '6-Mar', 'MARH6', prot$gene_symbol)

# Find values 3 SD above the mean of all CR values as cutoff
v <- mean(unlist(prot[,30:31])) + 3*(sd(unlist(prot[,30:31])))

# Filter for significant proteins
prot$crmax <- apply(prot[, 30:31], 1, max)
prot %<>% dplyr::filter(cr_reuterin_50 > v | cr_reuterin_100 > v)

# Select cols of interest
prot %<>% dplyr::select(gene_symbol, crmax)
colnames(prot) <- c('gene', 'crmax')

# take only highest cp value for each gene
prot %<>%
    group_by(gene) %>%
    summarise(crmax= max(crmax))

hold <- prot

```

### Add subsequent proteomics data
Analysis carried out in same method as the first analysis using mean+3SD cutoff. Ultimately if a gene is significant in either the initial or subsequent proteoimcs analysis it is counted as a hit. Only the highest CRmax value is taken for each gene. 
```{r}

# load and rename cols
prot <- fread('~/Desktop/projects/reuterin/data/UMichigan_Hannah_HCT_DMF_Reut_Cys_SLC-ABPP.txt', data.table = F)
colnames(prot)[c(2,28:32)] <- c('gene_symbol', 'cr_dmf_25', 'cr_dmf_100', 'cr_reuterin_50', 'cr_reuterin_100', 'gene_site')

# Correct excel's goofy date switches
prot$gene_symbol <- ifelse(prot$gene_symbol == '10-Sep', 'SEP10', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '9-Sep', 'SEPT9', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '11-Sep', 'SEP11', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '2-Sep', 'SEPT2', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '5-Sep', 'SEPT5', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '7-Sep', 'SEPT7', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '8-Sep', 'SEPT8', prot$gene_symbol)

prot$gene_symbol <- ifelse(prot$gene_symbol == '2-Mar', 'MOSC2', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '5-Mar', 'MARH5', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '7-Mar', 'MARH7', prot$gene_symbol)
prot$gene_symbol <- ifelse(prot$gene_symbol == '6-Mar', 'MARH6', prot$gene_symbol)

# Find values 3 SD above the mean of all CR values as cutoff
v <- mean(unlist(prot[,30:31])) + 3*(sd(unlist(prot[,30:31])))

# Filter for significant proteins
prot$crmax <- apply(prot[, 30:31], 1, max)
prot %<>% dplyr::filter(cr_reuterin_50 > v | cr_reuterin_100 > v)

# Select cols of interest
prot %<>% dplyr::select(gene_symbol, crmax)
colnames(prot) <- c('gene', 'crmax')

# take only highest cp value for each gene
prot %<>%
    group_by(gene) %>%
    summarise(crmax= max(crmax))

# combine new cysteine proteomics with old
prot <- rbind(hold, prot)

# take only highest cp value for each gene
prot %<>%
    group_by(gene) %>%
    summarise(crmax= max(crmax))

# combine with metabolite data
export <- merge(export, prot, all.x = T, by = 'gene')

# Add boolean proteomiccs value
export$cpBoolean <- ifelse(is.na(export$crmax), F, T)


```



# Convert Human metabolite/cysteiene proteomics to mouse 
RNA-seq was done in mouse, while the proteomics and metabolomics were done in human cell lines. 
```{r}
require("biomaRt")

# access biomart 
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")

# convert human common -> human ensembl -> mouse mgi
genes = getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = export$gene , mart = human, attributesL = c("ensembl_gene_id"), martL = human, uniqueRows=T)
colnames(genes) <- c('gene', 'ensembl')
genes2 = getLDS(attributes = c("ensembl_gene_id"), filters = "ensembl_gene_id", values = genes$ensembl , mart = human, attributesL = c("mgi_symbol"), martL = mouse, uniqueRows=T)
colnames(genes2) <- c('ensembl', 'mgi')

# choose single mgi id when returned 1+ for multiple human ensebml ids. Already optimally ordered for duplicate removal
t <- merge(genes, genes2, all.x = T, by = 'ensembl')
t <- t[!duplicated(t$gene),]

# merge new mouse gene names with pertinent metabolite/proteomics data and select cols
export <- merge(export, t, all.x = T, by='gene')
export %<>% dplyr::select(compound, interaction, mgi, crmax, cpBoolean) %>% unique()

```



# Add Gene Expression 
Add in RNA-seq data and export file for use with cytoscape. This export contains all the nodes but ones following will subset based on interesting compounds.
```{r}

# load rna
rna <- fread('~/Desktop/projects/reuterin/data/ReutvsCtrl.txt')

# get DEGs and select needed cols
rna %<>% dplyr::filter(FDR < 0.05) %>%
  dplyr::select(genes, logFC)

# select cols
colnames(rna) <- c('mgi', 'logFC')

# merge data frames and order cols
export <- merge(export, rna, all.x = T, by= 'mgi')

# Add differentially expressed column
export$de <- 100
export$de <- ifelse(export$logFC > 0, 1, export$de)
export$de <- ifelse(export$logFC < 0, -1, export$de)
export$de <- ifelse(is.na(export$logFC), 0, export$de)

# select desired cols
export %<>% dplyr::select(compound, interaction, mgi, crmax, cpBoolean, logFC, de)

# Add source target info
export$source <- T
export$target <- T

# Rename compounds for cytoscape viewing
export %<>%
  mutate(compound = ifelse(compound == 'OXIDIZED GLUATHIONE', 'GSSG', compound)) %>%
  mutate(compound = ifelse(compound == 'PHOSPHO-ENOL-PYRUVATE', 'PEP', compound)) %>%
  

# save info keeping only unique nodes
write.table(export, '~/Desktop/projects/reuterin/int/metabolite-gene.txt', sep = '\t', quote = F, row.names = F)


```


# Barplot of metabolite hits
Use this barplot to justify why the network was further narrowed down. Will show the number of DE genes as well as those with cysteine proteomics hits. 
```{r, fig.height=6, fig.width=5}

# create seperate data frame
plot <- export

# determine whether this gene-compound interaction has a DE gene
plot$total <- ifelse(plot$de != 0, 1, 0)

# Prepare for plotting
plot <- plot %>% 
  ##for each compound
  group_by(compound) %>% 
  ## get number of total genes associated
  mutate(numGenes = sum(total)) %>%
  ##calculate the proportion of the total genes DE
  mutate(proportion = sum(total)/length(compound)) %>%
  ##create a T/F value if gene is a cysteine proteomics hit
  mutate(cp = sum(cpBoolean)>0) %>%
  ##select cols and get unique vals
  dplyr::select(compound, numGenes, proportion, cp) %>%
  unique() %>%
  # arrange by number of genes and make factor for plotting
  arrange(numGenes) %>%
  mutate(compound = factor(compound, levels = .$compound)) %>%
  ## remove comounds with too few genes
  filter(numGenes > 0)
  
# Plot dotplot

g <- ggplot(plot, aes(numGenes, compound)) +
  geom_point(aes(size=proportion, col=cp)) +
  scale_color_manual(values=c("grey", "#00A08A")) +
  ggtitle(NULL) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(size="Proportion", colour= "CP hit") + 
  ylab(NULL) +
  xlab(NULL) +
  theme_minimal_grid(12)
g

pdf('~/Desktop/projects/reuterin/int/omicsIntegration_dotplot.pdf', height = 4, width = 5)
g
dev.off()

  

```

# Subset cytoscape export for only top hits
Choosing those with top # DE genes or any with a cysteine proteomics hit
```{r}

# keep the top x compounds and those with cystine proteomics hits
plot %<>% arrange(-numGenes)
## pull compund names
keep <- c(as.character(plot$compound[1:4]), as.character(plot[plot$cp >0,]$compound))
## get compounds of interest and select only the genes that have proteomics hit or are DE
export.sub <- export %>%
  dplyr::filter(compound %in% keep) %>%
  dplyr::filter(de != 0 | cpBoolean ==T)

# save info keeping only unique nodes
write.table(export.sub, '~/Desktop/projects/reuterin/int/metabolite-gene_subset.txt', sep = '\t', quote = F, row.names = F)


```


